---
title: "ESM 244 Lab 1"
author: "Ray Hunter"
date: "2023-01-13"
output: html_document
---

```{r setup, include=TRUE, echo = TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(here)
library(sf)
library(tmap)
```

order of operation to get info onto github
1) commit 
  go to "Git" tab and press commit
2) Pull
  blue down arrow - pulls info from online repository that may have been added since you worked on it. brings it to you local computer
3) Push
  puhes data back to github 
  
  
  in terminal 
  ls - tells you whats in the repository 
  
  
  
  
  
  ### Read in the data
```{r}
sf_trees <- read_csv(here('data', 'sf_trees', 'sf_trees.csv'))

#new data frame with just top 5 legal status trees 
top_5_status <- sf_trees %>% 
  group_by(legal_status) %>%
  summarize(tree_count=n()) %>%
  #finds the maximum values of the variable tree_count and slice out just the top 5
  slice_max(tree_count, n=5) %>%
  arrange(-tree_count)

ggplot(data = top_5_status, aes(x= fct_reorder(legal_status, tree_count), y = tree_count)) + 
#re order into how many trees are in eahc one 
  geom_col() +
  labs( y = "Tree Count ", x = "Legal Status") + 
 #swap the axis 
   coord_flip() +
theme_minimal()
```
  
in console, summary(sf_trees) 
basic summary of data

names(sf_trees)
shows the names of the variables

in console, 
sf_trees %>% group_by(legal_status) %>% summarize(tree_count = n())

shows how many observations there are for each category within a variable 
  
  
### filtering: keep observations where legal status is "permitted site" an dcare taker is "MTA"
```{r}
permitted_mta <- sf_trees %>% 
  filter(legal_status == "Permitted Sites" & caretaker == "MTA")
```


### Keep blackwood acacia trees, only lega status, date, lattitude, longitude colomns
```{r}
blackwood_acacia <- sf_trees %>%
  #str_detect looks for words withion a colomn even if there are other words
  filter(str_detect(species, "Blackwood Acacia")) %>%
  select(legal_status, date, latitude, longitude )

### make a plot 

ggplot(data = blackwood_acaia, aes(x = longitude, y = latitude)) +
  geom_point()
```

### use tidyr :: separate() to separate one colomn into multiple 
```{r}
sf_trees_separate <- sf_trees %>%
  separate(species, into = c("spp_scientific", "species_common"), sep = ' :: ')
```

### use tidy::unite to rejoin multiple colomns 
```{r}
sf_trees_unite <- sf_trees %>%
  unite("id status", tree_id:species, sep="_NEW_")
```


### Make some actual maps !

```{r}
blackwood_acacia_sf <- blackwood_acacia %>%
  drop_na(longitude, latitude) %>%
  #creating simple features spatial data creating a geometry point
  st_as_sf(coords = c("longitude", "latitude"))
#indicating that lattitude and longitude are degrees and not meters or feet
st_crs(blackwood_acacia_sf) <- 4326

ggplot(data = blackwood_acacia_sf) +
  #refers to the geometry colomn to choose the aesthetics of a map b/c it knows that we are talking about lat and long
  geom_sf(color= "darkgreen") + 
  theme_minimal()

```

### read in SF streets data 
```{r}
sf_map_sf <- read_sf(here("data", "sf_map", "tl_2017_06075_roads.shp")) %>%
  # transfor 4269 coordinates reference number to 4326
  st_transform(4326)
st_crs(sf_map_sf)
```



### MAPPING IT OUT 
```{r}
ggplot() + 
  geom_sf(data =sf_map_sf, size = 0.1, color = "darkgrey") + 
  geom_sf(data = blackwood_acacia_sf, color = "red", size = 0.5) + 
  #gets rid of axis
  theme_void() + 
  labs(title= "Blackwood Acacias in San Francisco")
```

  
### Interactive map 
```{r}
tmap_mode("view")
tm_shape(blackwood_acacia_sf) +
  tm_dots()
```

  
  
  
  
  
  
